Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-13 analysis/paper/fig_tab.Rmd
AGH final comments on first draft

---
title: "Tables and Figures"
output:
  bookdown::pdf_document2:
    toc: no
---

```{r setup-figs-and-tbls, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
#code to include R scripts in .Rmd: r knitr::spin_child("tables.R")
```

@@

@@

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorported into the samples.")
```

@@

```{r sep-plot, fig.cap="(A) Proportion (%) of sizes of starch granules in the wheat solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and (B) in the potato solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
pl_wheat1 + pl_potato1 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r mix-plot, fig.cap="Proportion of sizes of (A) wheat granules in the mixed solution (outer ring) and extracted from the wheat-treatment samples (inner ring), and sizes of (B) potato granules in the solution (outer ring) and extracted from the potato-treatment samples (inner ring). l = large, m = medium, s = small."}
#plot_grid(pl_wheat2, pl_potato2, labels = c("C", "D"), label_y = 0.1)
pl_wheat2 + pl_potato2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for seprated treatments."}
pl_cor
```

>>>>

---
title: "Tables and Figures"
output:
  bookdown::pdf_document2:
    toc: no
  bookdown::html_document2:
    toc: no
---

```{r setup-figs-and-tbls, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE)
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
```

@@

```{r anova-tbl}
knitr::kable(anova_tbl, 
             col.names = c("Treatment", "Mean", "SD", "Min", "Max"),
             caption = "Summary statistics for biofilm dry-weights (in mg) by treatment.")
```

```{r sample-count}
count_sample <- count_sample %>%
  ungroup() %>%
  add_row(count_sample_comb, .after = 2)

# convert columns to contain mean count and percentage of total count
count_samp_per <- count_sample %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), ")")))

count_samp_out <- tibble(count_samp_per, count_sample_sd, .name_repair = "unique") %>%
  select(1,2,3,7,4,8,5,9,6,10) #%>%

count_samp_out[1,3] <- NA # replace with NA
  #add_row(tibble_row(count_sample_comb_out), .before = 3)
```

```{r solution-count-out}
count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

### adding the 'both' row to the data frame ###
count_sol_comb_out <- count_sol_comb %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

count_sol_out <- count_sol %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")"))) %>%
  ungroup() %>%
  add_row(tibble_row(count_sol_comb_out), .before = 3)

count_sol_out[1,3] <- NA
```

```{r solution-count-tbl}
cnames <- c("Solution", "Starch", "Small (%)", "Medium (%)", "Large (%)", "Total (%)")
options(knitr.kable.NA = "")
knitr::kable(count_sol_out, col.names = cnames, 
             caption = "Mean starch counts from solutions, including the
             proportional makeup of the different sizes of granules.")
```

```{r sample-count-tbl}
cnames <- c("Treatment", "Starch", "Small (%)", "SD", "Medium (%)", "SD", 
            "Large (%)", "SD", "Total (%)", "SD")
knitr::kable(count_samp_out, col.names = cnames, 
             caption = "Mean starch counts extracted from samples with standard 
             deviation (SD), including the proportion of granule sizes of the total count.")
```

@@

```{r sample-prop-tbl}
cnames <- c("Treatment", "Starch", "Small", "Medium", "Large", "Total")
knitr::kable(perc, col.names = cnames, 
             caption = "The mean percentage of starches from the solutions that 
              were incorporated into the samples.")
```

@@

```{r protocol-fig, fig.cap="Overview of experiment protocol including the plate setup."}
knitr::include_graphics(here("analysis/figures/protocol_overview.png"))
```

```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\\mu$m.", out.width="50%"}
knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
```

```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
pl_cor
#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

### 2021-10-13 analysis/paper/paper.Rmd
AGH final comments on first draft

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored the how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs (substrata), which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins.

@@

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours, and no 
food was consumed two hours prior to donation. No antibiotics were taken up to six 
months prior to donation. 
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l Urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol).

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment, were included to detect starch contamination from the
environment, or cross-contamination from other wells in the same plate.

@@

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. This is a
reasonable assumption to make based on the small granule counts from the separate
treatment solutions, where the wheat granules made up 
`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
of all small granule counts in the separated solutions.
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

@@

```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments.", out.width="80%"}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments.", out.width="80%"}
pl_cor
#pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus.", out.width="80%"}
pl_cor2
```

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in 
diameter were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2)).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.

@@

tooth surfaces (including plaque and dental calulus) and mucous membranes 

@@

<!-- limitations of the platform show which questions would be unanswered -->
A limitation of this study 
The absence of $\alpha$-amylase in the model is a limitation of this study, as 
the total granule counts were not subject to hydrolysis. This would likely have 
reduced and affected the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
the absence can also allow us to directly explore
the effect of $\alpha$-amylase on starch counts in future experiments,
where $\alpha$-amylase can be added to the model
in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogeneous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on the comparatively low counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.

>>>>

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodMultiplePathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved when studying living organisms. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

@@

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationDentalPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

@@

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours. No 
food was consumed two hours prior to donation and no antibiotics were taken up 
to six months prior to donation. 
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol).

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment were included to detect starch contamination from the
environment or cross-contamination from other wells in the same plate.

@@

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this 
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the total amount of small starches in these two solutions, 
`r paste0(signif((count_sol$s[4] / sum(c(count_sol$s[3], count_sol$s[4]))) * 100, 3), "%")`
are from wheat.
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

@@

```{r ratio-plots, fig.cap="Proportion of sizes of starch granules from solutions (outer ring) and treatment samples (inner ring) in separated wheat (A) and potato (B) treatments, and mixed wheat (C) and potato (D) treatments."}
#plot_grid(pl_wheat1, pl_potato1, labels = c("A", "B"), label_y = 0.1)
(pl_wheat1 + pl_potato1) / (pl_wheat2 + pl_potato2) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

```{r cor-plot, warning=FALSE, fig.cap="Scatter plot of sample weight and standardised starch count by Z-score for separated treatments."}
#pl_cor
pl_cor + pl_cor2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

@@

```{r cor-plot2, warning=FALSE, fig.cap="Scatter plot of sample weight in mg and standardised count of starch grains per mg calculus."}
pl_cor2
```

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot));
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot2)).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.

@@

tooth surfaces (including plaque and dental calculus) and mucous membranes 

@@

<!-- limitations of the platform show which questions would be unanswered -->
The absence of $\alpha$-amylase in the model is a limitation of this study, as 
the total granule counts were not subject to hydrolysis. This would likely have 
reduced and affected the size ratios, as smaller starches may be more 
susceptible to hydrolysis 
[@francoStarchDegradation1992; @haslamDecompositionStarch2004]; however, 
the absence can also allow us to directly explore
the effect of $\alpha$-amylase on starch counts in future experiments,
where $\alpha$-amylase can be added to the model
in concentrations similar to those found in the oral cavity [@scannapiecoSalivaryAmylase1993]. 
We are able to show how absolute counts in the treatments cause a difference in
incorporation. However, this
was merely a side-effect of the difference in the number of granules in potato and 
wheat solutions of the same concentration (w/v). Further research should test 
multiple differing concentrations of the same starch type.
The use of EDTA may also have affected counts. While previous studies have shown 
negligible morphological changes caused by exposure to EDTA 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
these studies have not considered changes to separate size
categories within starch types, and whether shifts in size ratios occur due to
exposure to the pre-treatment chemicals.
The total number of granules on a slide often exceeded a number that 
was feasible to count in a reasonable time period, so we calculated the total 
counts by extrapolating from three slide transects. 
Thus, we reasonably assume that the three transects are a good representation 
of the entire slide, and that the distribution of all granules on the slide is 
relatively homogeneous.   
Finally, we only used native starches in the experimental procedure and the results 
will likely differ for processed starches [@graneroStarchTaphonomy2020]. 
Based on the comparatively low counts obtained by Leonard
and colleagues [-@leonardDentalCalculus2015, Supplement 2], processing and amylase
may have a substantial effect on starch granule retention in the oral cavity.
